\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-08 leslieandkenny.Rnw
worked on Model Refinement and Model Interpretation Sections

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}

@@

\begin{document}
<<setup, include=FALSE, cache=FALSE,echo=FALSE>>=
require(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, out.width = '\\linewidth',
               out.height = '0.4\\linewidth', dev = 'pdf',
               concordance = TRUE,
               tidy.opts = list(width.cutoff = 60, keep.blank.line=FALSE),
               size = 'footnotesize')
options(replace.assign = TRUE, width = 112, digits = 3, max.print = '72',
        show.signif.stars = FALSE)

@@

<<data,echo=FALSE>>=
# Read and set up the data
fert2 <- read.csv('fert2.csv')

@@

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots are independent given the other variables in the model.

@@

\noindent Before fitting a model, we recommend starting by creating plots
for exploratory data analysis. This is useful for seeing which interactions
may be present, and for identifying patterns that may represent violations
of assumptions or potential computational problems.

@@

\noindent We noticed that infection counts of 0 were recorded in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero.

\noindent We recommend sorting the data and creating a heatmap to look
for patterns of zeros. Figure 2 shows an example using simulated data.
The most serious problems will come from the zeros, so we chose a contrasting
color (black) for plots with infection counts of 0.

\noindent Any treatment or variety with many black squares is a reason
for concern. In the example, all but one of the inoculated fall-application
plots for Var1 had counts of 0. The software will have difficulty estimating
the the three-way interaction involving fall nitrogen application and
variety 1.

@@

\subsection{Model Fitting and Possible Issues}

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

@@

\begin{figure}
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
require(effects)
plot(Effect(c('variety','n.trt','status2'), glm3way),
     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety/Treatment/Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\end{figure}

@@

Recommend to use Chi Squared tests or F-tests? I think F-tests but not exactly sure what the F-test is giving or where its coming from.

Correct df? different from lme and glmer...

<<nosnms3way, echo=FALSE>>=
Anova(noSNMS3way)
anova(noSNMS3way)
#1-pf(2.29, 6, 118)

AIC(noSNMS3way, noSNMS2way)
anova(noSNMS3way, noSNMS2way)

#lme model to get correct df for f tests
#why is den df now 100?
require(nlme, quietly=T)
noSNMS3waylme <- lme(infected ~ variety*n.trt*status,
                    random=~1|block/row, method="ML",
                     data = subset(fert2, variety != "SNMS"))

glm3way <- glm(cbind(infected, total) ~ variety*n.trt*status+row, family="binomial", data=subset(fert2, variety != "SNMS"))
@

explain: if no evidence of a three way interaction what can you do, how to continue selection.

Adding a sentence to make sure I can push

@@

Explain how to interpret with three way interaction (essentially interpret two way interaction for each inoculated and control)

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{float}

@@

\begin{document}
<<setup, include=FALSE, cache=FALSE,echo=FALSE>>=
require(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, out.width = '\\linewidth',
               out.height = '0.4\\linewidth', dev = 'pdf',
               concordance = TRUE,
               tidy.opts = list(width.cutoff = 60, keep.blank.line=FALSE),
               size = 'footnotesize')
options(replace.assign = TRUE, width = 112, digits = 3, max.print = '202',
        show.signif.stars = FALSE)

@@

<<data,echo=FALSE>>=
# Read and set up the data
#for leslie 
#setwd("~/Documents/ConsultingFall2015/Fall510")
fert2 <- read.csv("fert2.csv")

@@

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots are independent given the other variables in the model. \\

@@

\noindent Before fitting a model, we recommend starting by creating plots
for exploratory data analysis. This is useful for seeing which interactions
may be present, and for identifying patterns that may represent violations
of assumptions or potential computational problems. \\

@@

\noindent We noticed that infection counts of 0 were recorded in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero. \\

\noindent We recommend sorting the data and creating a heatmap to look
for patterns of zeros. Figure 2 shows an example using simulated data.
The most serious problems will come from the zeros, so we chose a contrasting
color (black) for plots with infection counts of 0.\\

\noindent Any treatment or variety with many black squares is a reason
for concern. In the example, all but one of the inoculated fall-application
plots for Var1 had counts of 0. The software will have difficulty estimating
the the three-way interaction involving fall nitrogen application and
variety 1. \\

@@

\subsection{Model Fitting and Possible Issues}
\label{model}

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
#this forces the constraint that the block effects sum to 0
contrasts(fert.sim1$block) <- contr.sum(6, contrast=TRUE)
glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

@@

@@

In this section, we describe how to go through the process of model selection. As discussed in the previous section, we recommend starting with the most complicated $3$ way interaction model and then assessing which terms can be removed. In general, start by considering removal of the highest order terms. \\

\noindent We'll start by assessing evidence for the three-way interaction term. One way to consider it's removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$. This is method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package, and choosing the option \verb+test="F"+. This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals (provide reference here to info about pearson resids vs deviance resids). This terms in this table are tested conditional upon terms of the same and lower order being in the model. For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model. As a result, this table can be used to assess evidence for removal of each term.

<<refine, echo=TRUE>>=
Anova(glm3way, test="F")
@

\noindent Here is how we would interpret the results in each row of the table. Note, it is not appropriate to consider removal of any of the terms involving block. Block to block variability is expected because it was included as a design variable, and so it should be included in the model. The $block*variety$ interaction also needs to be included to account for the nesting structure in the design. 
\begin{enumerate}
\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time depends on inoculation status (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.368$ from F-stat$=1.01$ on $2$ and $133$ df).
\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.288$ from F-stat$=1.26$ on $4$ and $133$ df).
\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions in the model (p-value$=0.453$ from F-stat$=0.98$ on $8$ and $133$ df).
\end{enumerate}

\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (shown below). If this model is selected, each effect is then interpreted after controlling for the effects of the other variables in the model. For example, to assess evidence for the \verb+n.trt+ effect in this model, we would say, ``There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the 
$block*variety$ interaction.''

<<glmnointeractions, echo=TRUE>>=
glmnointeractions <- glm(cbind(infected, total) ~ block/variety + variety+n.trt 
                         + status2,
               family = quasibinomial, data = fert.sim1)
Anova(glmnointeractions, test="F")
@

\noindent The next section describes how to interpret effects once a model is selected.

<<dropindevftest, echo=FALSE>>=
glm.no3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + n.trt*status2 + variety*status2,
               family = quasibinomial, data = fert.sim1)
#drop in dev F-test
#((196-185)/8)/(185/125) 
#0.929
#1-pf(0.929, 11, 125)
#0.512
@

@@

\noindent In this section, we describe how to interpret effects if evidence of a three-way interaction is found. The effects plots in the \verb+effects+ packages provide nice visuals for displaying estimated probabilities of infection across levels of mulitiple variables.

\begin{figure}[!h]
<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5, fig.align='center'>>=
require(effects)
plot(Effect(c('variety','n.trt','status2'), glm3way),
     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
     main = 'Variety/Treatment/Inoculation Effects Plot',
     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
     key.args = list(title = 'Inoculation Status'))
@
\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
\end{figure}

\noindent The \verb+Effect+ function is convenient because it provides the estimated probabilities at each treatment combination, averaged over all blocks. The results are  provided in a organized table (see below).

<<summary, echo=TRUE>>=
Effect(c('variety','n.trt','status2'), glm3way, se=TRUE)
@

\noindent We recommend displaying the effects plot and output and then describing the effects of interest in words. When describing the effects in words, we recommend describing the effects in terms of multiplicative changes in the odds of virus infection. It is more straightforward to calculate confidence intervals when effects are described are this scale rather than the probability scale. We recommend changing the reference level of the model to find the estimates and confidence intervals for the effects of interest. Notice that when we fit the model in Section \ref{model}, we specified the constraint that the sum of the block effects would add to $0$. As a result, the estimated effects given in the model summary are the effects averaged over all blocks. \\

\noindent You can start by looking at the summary of the model, \verb+summary(glm3way)+. The coefficients of interest will be those found in the rows \verb+n.trtEARLY SPRING+ and \verb+n.trtLATE SPRING+. Below we show how to interpret these coefficients in words, and how the meaning changes when the reference level is manipulated.

<<first, echo=TRUE, message=FALSE>>=
#use the default reference level to describe the effects across 
#nitrogen application for inoculated plants of variety 1 (averaged over blocks)
exp(coef(glm3way)[11])
exp(confint(glm3way, 11))
exp(coef(glm3way)[12])
exp(confint(glm3way, 12))
@

\noindent For these data, we would describe the nitrogen application effect for inoculated plants of variety $1$ as follows:
\begin{quote}
For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $3.01$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.295$ to $104.9$ times. For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $5.05$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.652$ to $166.4$ times.
\end{quote}

<<second, echo=TRUE, message=FALSE, eval=FALSE>>=
#suppose we want to the effects of introgen application for 
#inoculated plants of variety 3
#first relevel
var3 <- relevel(fert.sim1$variety, "Var3")
#then fit model with var3
glm.refvar3 <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
#interpret results
exp(coef(glm.refvar3)[11])
exp(confint(glm.refvar3, 11))
exp(coef(glm.refvar3)[12])
exp(confint(glm.refvar3, 12))

#now suppose we want to the effects of introgen application for 
#non-inoculated plants of variety 3
#first relevel the status variable
statuscntl <- relevel(fert.sim1$status2, "CNTL")
#then fit model with var3
glm.refvar3cntl <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*statuscntl,
               family = quasibinomial, data = fert.sim1)
#interpret results
exp(coef(glm.refvar3cntl)[11])
exp(confint(glm.refvar3cntl, 11))
exp(coef(glm.refvar3cntl)[12])
exp(confint(glm.refvar3cntl, 12))
@

\noindent We would describe the nitrogen application effect for inoculated plants of variety $3$ as follows:
\begin{quote}
For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be about equal to the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.191$ to $5.28$ times. For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.24$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.269$ to $6.201$ times.
\end{quote}

\noindent We would describe the nitrogen application effect for non-inoculated plants of variety $3$ as follows:
\begin{quote}
For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $2.3$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.797$ to $7.72$ times. For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.45$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.446$ to $5.16$ times.
\end{quote}
